### 2020-08-28 serial.tex
tweaks

The total density of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at time \psymp, relative to the total number of serial intervals starting at time \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}.
\label{eq:fserial}
\end{equation}
The denominator, $\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'$, then corresponds to the total number of symptomatic cases generated by primary cases who developed at time \psymp;
dividing it by the incidence of symptomatic cases (also infected by symptomatic cases) at time \psymp, $j(\psymp) = \int_{-\infty}^\infty \total(\psymp-\tau', \psymp) \dtau'$, we obtain the serial reproduction number:
\begin{equation}
\Rs(\psymp) = \frac{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}{j(\psymp)},
\end{equation}
which we define as the average number of symptomatic cases generated by individuals who developed symptoms at time \psymp (and were also infected by sympmtomatic cases).
Combining the forward serial-interval distribution with the serial reproduction number allows us to express the renewal process between symptomatic cases:
\begin{equation}
j(t) = \int_{-\infty}^\infty \Rs(t-\tau) j(t-\tau) f_{t-\tau}(\tau) \dtau.
\label{eq:jrenew}
\end{equation}
In Supplementary Materials, we show that estimating the serial reproduction number requires the backward serial-interval distribution $b_\ssymp(\tau)$, which is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int_{-\infty}^\infty \total(\ssymp-\tau', \ssymp) \dtau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any epidemic model and link serial-interval distributions with the renewal process.

@@

The initial forward serial-interval distribution depends on the exponential growth rate $r$.
For a fast-growing epidemic (high $r$), we expect the backward incubation periods to be short (\eref{backexp}), meaning that presymptomatic transmission is less likely to occur; 
therefore, the initial forward serial-interval distribution will generally have a larger mean than the intrinsic generation- and serial-interval distributions.
However, the exact shape of the initial forward serial-interval distribution depends on the shape of the joint distribution.
For example, the Susceptible-Exposed-Infected-Recovered model, under the additional assumption that the incubation and exposed periods are equivalent (i.e. that onset of symptoms and infectiousness occur simultaneously), provides a special case where the forward serial- and generation-intervals follow the same distributions during the exponential growth phase because (i) infected individuals can only transmit after symptom onset and (ii) the time between symptom onset and infection is independent of the incubation period of an infector (see Supplementary Materials).
Throughout the paper, we allow for presymptomatic transmission to match the transmission dynamics of COVID-19.a

@@

The dynamics of the serial-interval distribution depends heavily on the joint distribution between incubation periods and generation intervals.
Here, we use a bivariate lognormal distribution to model the joint probability distribution of the intrinsic incubation periods and the intrinsic generation intervals while allowing for the possibility that they might be correlated.
Given that the viral load of SARS-CoV-2 peaks around the time of symptom onset, we generally expect the generation intervals to be positively correlated with the incubation period;
that is, individuals who develop symptoms later are more likely to transmit later.
Marginal distributions of incubation periods and generation intervals are parameterized based on parameter estimates for COVID-19 (Table 1).
For simplicity, we consider three values for the correlation coefficients (on the log scale) of the bivariate lognormal distribution: $\rho = -0.5, 0, 0.5$.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
  \textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The intrinsic incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a bivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
The intrinsic generation-interval and incubation period distributions are chosen to match characteristic of COVID-19 to illustrate realistic magnitudes of time-varying/perspective effects in the current pandemic.
}
\end{table}

@@

The initial forward serial-interval distribution captures the exponential growth phase of an epidemic.
We explore how forward and backward serial intervals can vary over the course of an epidemic using deterministic and stochastic simulations based on the renewal equations (see Supplementary Materials) using parameters in Table 1;
we further assume $\Ro=2.5$.
While the forward serial-interval distribution is our primary focus, understanding the differences between the forward and the backward distributions is important because the observed intervals during an ongoing epidemic are often the backward ones:
we typically identify infected individuals and ask when and by whom they were infected.
Similarly, when we are estimating the incubation period of an individual, we typically observe their symptom onset date and try to estimate when they were infected (e.g., \cite{backer2020incubation}).

>>>>

The total density of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at time \psymp, relative to the total number of serial intervals starting at time \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}.
\label{eq:fserial}
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int_{-\infty}^\infty \total(\ssymp-\tau', \ssymp) \dtau'}.
\label{eq:bserial}
\end{equation}

We see that the denominator of the forward serial-interval distribution, $\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'$, then corresponds to the total number of symptomatic cases generated by primary cases who developed at time \psymp (\eref{fserial});
dividing it by the incidence of symptomatic cases (also infected by symptomatic cases) at time \psymp, $j(\psymp) = \int_{-\infty}^\infty \total(\psymp-\tau', \psymp) \dtau'$, we obtain the serial reproduction number:
\begin{equation}
\Rs(\psymp) = \frac{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}{j(\psymp)},
\end{equation}
which we define as the average number of symptomatic cases generated by individuals who developed symptoms at time \psymp (and were also infected by sympmtomatic cases).
Combining the forward serial-interval distribution with the serial reproduction number completes the renewal process between symptomatic cases:
\begin{equation}
j(t) = \int_{-\infty}^\infty \Rs(t-\tau) j(t-\tau) f_{t-\tau}(\tau) \dtau.
\label{eq:jrenew}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any epidemic model and properly link serial-interval distributions with the renewal process.

@@

The initial forward serial-interval distribution depends on the exponential growth rate $r$.
For a fast-growing epidemic (high $r$), we expect the backward incubation periods to be short (\eref{backexp}), meaning that presymptomatic transmission is less likely to occur; 
therefore, the initial forward serial-interval distribution will generally have a larger mean than the intrinsic generation- and serial-interval distributions.
However, the exact shape of the initial forward serial-interval distribution depends on the shape of the joint distribution.
For example, the Susceptible-Exposed-Infected-Recovered model, under the additional assumption that the incubation and exposed periods are equivalent (i.e. that onset of symptoms and infectiousness occur simultaneously), provides a special case where the forward serial- and generation-intervals follow the same distributions during the exponential growth phase because (i) infected individuals can only transmit after symptom onset and (ii) the time between symptom onset and infection is independent of the incubation period of an infector (see Supplementary Materials).
Throughout the paper, we allow for presymptomatic transmission to match the transmission dynamics of COVID-19.

@@

We have shown that the dynamics of the serial-interval distribution depends on the joint distribution between incubation periods and generation intervals.
Here, we use a bivariate lognormal distribution to model the joint probability distribution $h$ of intrinsic incubation periods and intrinsic generation intervals (in the renewal equation model, \eref{renewal}) while allowing for the possibility that they might be correlated.
Given that the viral load of SARS-CoV-2 peaks around the time of symptom onset, we generally expect the generation intervals to be positively correlated with the incubation period:
that is, individuals who develop symptoms later are more likely to transmit later.
Marginal distributions of incubation periods and generation intervals are parameterized based on parameter estimates for COVID-19 (Table 1).
For simplicity, we consider four values for the correlation coefficients (on the log scale) of the bivariate lognormal distribution: $\rho = 0, 0.25, 0.5, 0.75$.
This parameterization allows for generation intervals to be shorter than the incubation period, allowing for presymptomatic transmission.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
  \textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The intrinsic incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a bivariate log-normal distribution with correlations (on the log scale) $\rho=0, 0.25, 0.5, 0.75$.
The intrinsic generation-interval and incubation period distributions are chosen to match characteristic of COVID-19 to illustrate realistic magnitudes of time-varying/perspective effects in the current pandemic.
}
\end{table}

@@

The initial forward serial-interval distribution captures the exponential growth phase of an epidemic.
We now explore how forward and backward serial intervals can vary over the course of an epidemic using deterministic and stochastic simulations based on the renewal equations (see Supplementary Materials) using parameters in Table 1;
we further assume $\Ro=2.5$ to reflect the transmission dynamics of COVID-19 \citep{doi:10.1098/rsif.2020.0144}.
While the forward serial-interval distribution is our primary focus, understanding the differences between the forward and the backward distributions is important because the observed intervals during an ongoing epidemic are often the backward ones:
we typically identify infected individuals and ask when and by whom they were infected.
Similarly, when we are estimating the incubation period of an individual, we typically observe their symptom onset date and try to estimate when they were infected (e.g., \cite{backer2020incubation}).
